
```{r}
#read libraries
##general data handling
library(tidyverse)
library(dplyr)
library(reticulate)
library(stringr)
#spatial wrangling and plotting
library(leaflet)
library(sf)
library(mapview)
library(tmap)
#data reading/processing
library(readxl)
library(jsonlite)
library(purrr)

#statistical analysis
library(EnvStats)
library(cmdstanr)
library(rethinking)
```

If you need to install cmdstanr, use this, as it is not on CRAN:
```{r}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos"))) 
library(cmdstanr)
install_cmdstan()
```

#Section 1: Voting Areas

## load polygons and wrangle
from https://data-science.dk/datasat/gis/valg/valgsteder/afstemningsomraader-2026-fv26-download-de-danske-valgsteder/
```{r}
data <- read_sf("../Raw_data/fv2026.shp")

#check if it looks right
#plot(st_geometry(data))

#check crs and rename to keep track of it
st_crs(data) 
data_utm <- data

#make unique names by combining colloqiual name and voting area index
data_utm$full_name <- paste0(data_utm$navn, ";",data_utm$opstilling)

#sort by this name
data_utm <- data_utm[order(data_utm$full_name),]

#as we will later see, these two have the exact same polling station locations, so we will merge their polygons
which(data_utm$navn %in% c("Gymnasiet 2","Gymnasiet 1"))
data_utm[385,]$geometry <- st_union(data_utm[c(385,386),])
data_utm <- data_utm[c(-386),]

#also remove Christiansø, due to mismatch with BBR in section 4
data_utm <- data_utm[c(-which(data_utm$navn == "Christiansø")),]
```

#Section 2: Polling station locations

```{r}
data <- fromJSON("https://api.dataforsyningen.dk/afstemningsomraader", simplifyVector = FALSE)

polling_points <- map_dfr(data, function(x) {
  tibble(
    navn = x$navn,
    lng = x$afstemningssted$adgangsadresse$koordinater[[1]],
    lat = x$afstemningssted$adgangsadresse$koordinater[[2]],
    opstilling = x$opstillingskreds$nummer
  )
})


polling_points %>% select(c("lng","lat")) %>% head() #looks like wgs84

polling_points_wgs <- st_as_sf(polling_points, coords = c("lng", "lat"), crs = 4326)
polling_points_utm <- st_transform(polling_points_wgs, 25832)

#make unique names by combining colloqiual name and area index
polling_points_utm$full_name <- paste0(polling_points_utm$navn, ";", polling_points_utm$opstilling)

#sort
polling_points_utm <- polling_points_utm[order(polling_points_utm$full_name),]


polling_points_utm[polling_points_utm$navn %in% c("Gymnasiet 2","Gymnasiet 1"),]$geometry #xmin and xmax of bounding boxx are equal, same for y, so these points have same location (to this level of precision).

#remove one of them, and Christiansø
polling_points_utm <- polling_points_utm[c(-which(polling_points_utm$navn %in% c("Gymnasiet 2", "Christiansø"))),]
```

Check that polling points align with voting regions
```{r}
has_own_point <- c()
for (i in 1:nrow(data_utm)){
  has_own_point <- c(has_own_point, lengths(st_intersects(data_utm$geometry[order(data_utm$full_name)][i], polling_points_utm[order(polling_points_utm$full_name)[i],])))
}

sum(has_own_point==1) #almost all polling stations lie within corresponding voting area

#look at those that don't
mapview(data_utm[order(data_utm$full_name),][has_own_point==0,], col.regions = "red") + 
  mapview(polling_points_utm[order(polling_points_utm$full_name),][has_own_point==0,])
#nothing to worry about, looks realistic
```

## Make matrix with distances between stations
```{r}
#make the matrix
distance_matrix <- st_distance(polling_points_utm, polling_points_utm)

#take a look - unit is meter
distance_matrix[c(1:5, 100:105),1:5]

#save as numeric values for easier computation
numeric_distance_matrix <- drop_units(distance_matrix)
numeric_distance_matrix <- numeric_distance_matrix/1000 #use km as units
```

#Section 3: Distribution of votes for each party and voter tournout for each voting place
From https://www.statistikbanken.dk/20005
```{r}
total_voting <- read_excel("../Raw_data/stor2026426145511622605080FV26TOTA.xlsx")

voters_data <- rep(0,1420) #make column to bind to
for (i in 0:15){
  relevant_votes <- total_voting[(3 + 1420*i):(2 + 1420*(i+1)),3]
  relevant_votes$...3 <- as.numeric(relevant_votes$...3)
  colnames(relevant_votes) <- total_voting[3 + 1420*i,1]
  voters_data <- cbind(voters_data, relevant_votes)
}
voters_data <- voters_data[,c(-1)]
voters_data$Område <- as.vector(total_voting[3:1422,2])[[1]]

colnames(voters_data)[c(1,15,16)] <-c("GYLDIGE_STEMMER_I_ALT", "UGYLDIGE_STEMMER_I_ALT", "ANTAL_VÆLGERE")

#get total count of votes
voters_data$Stemmer_I_Alt <- voters_data$GYLDIGE_STEMMER_I_ALT + voters_data$UGYLDIGE_STEMMER_I_ALT

#sanity check: sum of votes for parties and individuals equals sum of valid votes
sum(apply(voters_data[,c(2:14)],1,sum) - voters_data$GYLDIGE_STEMMER_I_ALT != 0)

#make "Område" the first column
voters_data <- voters_data %>% select(c(17,15, 1:14,16,18))

dim(total_voting)
```
### Correct voters_data
fix discrepancy between names in data_utm and voters_data:
all the ones with "og" have upper case "O" in data_utm but lowercase in voters_data.
```{r}
#Correct the names in voters_data
voters_data[voters_data[,1] == "Hedensted, St.Dalby og Urlev",1] <- "Hedensted, St.Dalby Og Urlev"

voters_data[voters_data[,1] == "Daugård og Ørum Sogne",1] <- "Daugård Og Ørum Sogne"

voters_data[voters_data[,1] == "Langå, Svindinge og Øksendrup",1] <- "Langå, Svindinge Og Øksendrup"
voters_data[voters_data[,1] == "Ølsted og Eriknauer-Området",1] <- "Ølsted Og Eriknauer-Området"
voters_data[voters_data[,1] == "Ellested og Herrested",1] <- "Ellested Og Herrested"
voters_data[voters_data[,1] == "Kullerup og Refsvindinge",1] <- "Kullerup Og Refsvindinge"
voters_data[voters_data[,1] == "Frørup og Tårup",1] <- "Frørup Og Tårup"

sum(!(data_utm$navn %in% voters_data[,1])) #none missing
```

Remove rows of voters_data that do not correspond to individual polling stations
```{r}
#first combine "Gymnasiet" 1 and 2, remove 2, and remove Christiansø
voters_data[voters_data$Område == "Gymnasiet 1",2:18] <- voters_data[voters_data$Område == "Gymnasiet 1",2:18] + voters_data[voters_data$Område == "Gymnasiet 2",2:18]
voters_data <- voters_data[c(-which(voters_data$Område == "Gymnasiet 2")),]
voters_data <- voters_data[c(-which(voters_data$Område == "Christiansø")),]

not_in_navn <- which(!(voters_data[,1] %in% data_utm$navn))
length(not_in_navn)
#106 left: these are overall area names, that do not correspond to individual polling stations: 1 "HELE LANDET" + 10 storkrede + 3 regioner + 92 opstillingskredse
#voters_data[not_in_navn,1] #check that names match the above description

#filter away
voters_voting_places_only <- which((voters_data[,1] %in% data_utm$navn))
reduced_voters <- voters_data[voters_voting_places_only,]

#make full_name for reduced_voters to have corresponding names
##reduced_voters data came ordered by opstillingskreds, and can be added like so:
reduced_voters$opstilling <- data_utm$opstilling[order(as.numeric(data_utm$opstilling))]

reduced_voters$full_name <- paste0(reduced_voters$Område, ";", reduced_voters$opstilling)

reduced_voters <- reduced_voters[order(reduced_voters$full_name),]
```

#Section 4: load BBR data
The file versions loaded when this analysis was first run may have expired, so you may need to use the below url (add you api key) to check
available file versions:

https://api.datafordeler.dk/FileDownloads/v2.0/GetAvailableFileDownloads?Register=BBR&apiKey=[api_key]

look/search for a filename of:
BBR_V3_Bygning_[4-digitmunicipality code]_TotalDownload_csv_Current_[version_number]
to find the newest version number available

## find distance from building in voting area to relevant polling station

### first a proof of concept to check that the approximation works
```{r}
#check that estimation works for finding mean p. It does.
b <- 4 #arbitrary value
estimates <- c()
errors <- c()
for (i in 1:1000){
  values <- rgamma(400, shape = 3, scale = 200) 
  est <- egamma(values, method = "mle")
  estimates <- rbind(estimates, est$parameters)
  
  true_mean <- mean(exp(-b*values))
  est_mean <- (1+est$parameters[2]*b)^(-est$parameters[1])
  error_mean <- true_mean-est_mean
  errors <- c(errors, error_mean)
} 

plot(estimates) #estimates of distribution shape are somewhat imprecise
hist(errors) #but errors are tiny, so its not a problem
quantile(errors)
```

### Now do the real thing
This loops through each municipality, loads all buildings from BBR in that municipality, extracts only buildings that people live in (from https://bbr.dk/kodelister):

Kode: 110 - Stuehus til landbrugsejendom
Kode: 120 - Fritliggende enfamiliehus
Kode: 121 - Sammenbygget enfamiliehus
Kode: 122 - Fritliggende enfamiliehus i tæt-lav bebyggelse
Kode: 130 - (UDFASES) Række-, kæde-, eller dobbelthus (lodret adskillelse mellem enhederne)
Kode: 131 - Række-, kæde- og klyngehus
Kode: 132 - Dobbelthus
Kode: 140 - Etagebolig-bygning, flerfamiliehus eller to-familiehus
Kode: 150 - Kollegium
Kode: 160 - Boligbygning til døgninstitution
Kode: 190 - Anden bygning til helårsbeboelse

...and extracts their coordinates. It then loops through all voting areas in that municipality, extracts residencies in that area, estimates the distributional parameters of the distances to the relevant polling station.

```{r}
#check that indices align
sum(data_utm$full_name != polling_points_utm$full_name)

api_key <- "write_key_here"

#make new columns to save distribution parameters
data_utm$k_estimates <- NA
data_utm$theta_estimates <- NA

outdir  <- tempdir()

kommunekoder <- unique(data_utm$kommunekod)

for(kommune in kommunekoder){
  
  # fetch zip file with data
  url <- paste0(
  "https://api.datafordeler.dk/FileDownloads/v1.0/GetFile?Filename=BBR_V3_Bygning_",kommune,"_TotalDownload_csv_Current_633.zip&apikey=",api_key
  )
  #read csv file in zip
  zipfile <- tempfile(fileext = ".zip")
  download.file(url, zipfile, mode = "wb")
  files <- unzip(zipfile, exdir = tempdir())
  temp_data <- read.csv(files[1])

  #filter to only get buildings that people live in
  bolig_data <- temp_data[temp_data$byg021BygningensAnvendelse %>% as.numeric() %in% c(110,120,121,122,130,131,132,140,150,160, 190),]

  #extract coordinates from data
  byg_lat <- c()
  byg_long <- c()
  strings <- bolig_data$byg404Koordinat %>% strsplit(" ")
  for (i in 1:length(strings)){
    byg_long[i] <- strings[[i]][2] %>% str_extract("\\d+\\.*\\d*")
    byg_lat[i] <- strings[[i]][3] %>% str_extract("\\d+\\.*\\d*")
  }

  #Convert to points
  byg_points_sf  <- st_as_sf(data.frame("long" = byg_long, "lat" = byg_lat), coords = c("long", "lat"), crs = 25832)
  byg_points_sf_saved <- byg_points_sf
  
  #go through each area to calculate distances
  
  for (area in which(data_utm$kommunekod == kommune)){
    
    inside <- st_within(
    byg_points_sf,
    data_utm[area,],
    sparse = FALSE
  )[,1]
    
    relevant_points <- byg_points_sf[inside,]
    
    #remove from list of buildings for future areas for faster search
    byg_points_sf <- byg_points_sf[!inside,]
    # define target
    polling_station <- polling_points_utm[area,]
    #save mean log distance for each area
    distances <- st_distance(polling_station, relevant_points) %>% drop_units() #drop meter unit. 
    estimates <- egamma(distances[1,], method = "mle")
    k_est <- estimates$parameters[1]
    theta_est <- estimates$parameters[2]
    
    data_utm$k_estimates[area] <- k_est
    data_utm$theta_estimates[area] <- theta_est
  }
}

#check if any data is missing
sum(is.na(data_utm$k_estimates))
sum(is.na(data_utm$theta_estimates))

hist(data_utm$k_estimates*data_utm$theta_estimates/1000, main = "Estimated mean distance to polling station for each area", xlab = "Distance (km)")
```

#Section 5: run stan models
## Prepare data
```{r}
n_voters <- reduced_voters$ANTAL_VÆLGERE

V <- reduced_voters$Stemmer_I_Alt

votes <- reduced_voters[,c(2,4:16)]
votes <- votes %>% as.matrix()

#make a prior for distribution of votes by using the national average
votes_share_pr_party <- apply(votes,2,mean)/sum(apply(votes,2,mean))
alpha_prior_mu <- log(vote_share_pr_party[1:13]/vote_share_pr_party[14])

votes_data_list <- list(
  N_regions = nrow(data_utm),
  N_parties = ncol(votes),
  votes = votes,
  Dmat = numeric_distance_matrix,
  alpha_mu = alpha_prior_mu
)



lost_votes_data_list <- list(
  V = V,
  n_voters = n_voters,
  k = data_utm$k_estimates,
  theta = data_utm$theta_estimates/1000, #adjust distances to be in km rather than m
  N_regions = nrow(data_utm),
  D = numeric_distance_matrix
)
```

## run models
This is very computationally heavy - consider running smaller version, or download the fitted values from the github repository
###votes_model
```{r}
#compile model and fit
votes_model <- cmdstan_model("votes_gp_1.stan")

votes_fit <- votes_model$sample(
  data = votes_data_list,
  chains = 10,
  parallel_chains = 10,
  iter_warmup = 400,
  iter_sampling = 200,
  adapt_delta = 0.90
)
```

###lost_votes_model
```{r}
lost_votes_model <- cmdstan_model("lost_votes.stan")

lost_votes_fit <- lost_votes_model$sample(
  data = lost_votes_data_list,
  chains = 20,
  parallel_chains = 20,
  iter_warmup = 400,
  iter_sampling = 100,
  adapt_delta = 0.90
)
```


#Section 6: analysing posterior draws

This section uses the posterior samples to simulate the outcome of a counterfactual election with 0 distance to polling stations for all voters, compared to simulated election with distances as they are today.

If you ran the models yourself:
```{r}
p_partes <- votes_fit$draws("p")
p_max <- lost_votes_fit$draws("p_max")
p_true <- lost_votes_fit$draws("p_true")
idx_used <- 1:1312 #change as needed
subset_data <- data_utm[idx_used,]
N_regions_used <- 1312
N_samples <- 2000 #adjust as needed: should be iter_sampling *chains
```

to read files from git:
```{r}
p_parties <- readRDS("../small_posterior_draws/p_parties_small.rds")
p_max <- readRDS("../small_posterior_draws/p_max_small.rds")
p_true <- readRDS("../small_posterior_draws/p_true_small.rds")
idx_used <- readRDS("../small_posterior_draws/idx_656_seed_123.rds")
subset_data <- data_utm[idx_used,]

N_regions_used <- 656
N_samples <- 300
```

Reshape p_max and p_true:
```{r}
#reshape arrays
max_proportions <- array_reshape(p_max, c(N_samples,N_regions_used), order="C")
true_proportions <- array_reshape(p_true, c(N_samples,N_regions_used), order="C")
#get contrast: how much higher would voter turnout be in each area 
contrast_proportions <- max_proportions-true_proportions
```

plot of proportion of voters lost:
```{r}
100*contrast_proportions %>% apply(2,HPDI,0.95) %>% range()#densest 95% credible interval posits that voter turnout in different areas would increase by at least 0.03 percentage points, and at most 12.76 percentage points

100*max(apply(contrast_proportions,2,mean)) #the highest posterior mean estimate of the difference in turnout for any area is 7.62 percentage points.

#Calculate the percentage point difference in voter turnout for all of denmark
extra_voters <- (contrast_proportions %*% diag(n_voters[idx_used])) %>% apply(1,sum)

100*(extra_voters/sum(n_voters[idx_used])) %>% HPDI(0.95) #95% credible interval from 0.76 to 1.81 percentage points. 


#make plot
subset_data$contrast_proportions_mu <- apply(contrast_proportions,2,mean)*100 #convert to percentage points

my_breaks <- c(0,1,2,4,8)

my_col <- rgb((my_breaks - my_breaks[1])/max(my_breaks - my_breaks[1]),1-(my_breaks - my_breaks[1])/max(my_breaks - my_breaks[1]),0)

tm_shape(subset_data) +
  tm_polygons(
    col = "contrast_proportions_mu",
    palette = my_col,
    breaks = my_breaks,
    title = "Percentage Point \nDifference in Turnout"
  ) +
  tm_layout(
    legend.outside = TRUE
  )
```

correct shape of p_parties array:
```{r}
p_party_re <- array_reshape(p_parties, c(N_samples,(N_regions_used*14)), order="C")
party_probabilities <- array_reshape(p_party_re, c(N_samples,N_regions_used, 14), order="F")
```

### Bias in outcome?
```{r}
sim_votes_true <- c()
for (draw in 1:2000){
  this_draw <- c()
  for (area in 1:N_regions_used){
    this_area <- rmultinom(1,n_voters[idx_used][area] * true_proportions[draw,area],party_probabilities[draw,area,])
    this_draw <- rbind(this_draw, t(this_area))
  }
  this_draw_flatten <- apply(this_draw,2,sum)/sum(this_draw)
  sim_votes_true <- rbind(sim_votes_true, this_draw_flatten)
}

sim_votes_max <- c()
for (draw in 1:N_samples){
  this_draw <- c()
  for (area in 1:N_regions_used){
    this_area <- rmultinom(1,n_voters[idx_used][area] * true_proportions[draw,area],party_probabilities[draw,area,])
    this_draw <- rbind(this_draw, t(this_area))
  }
  this_draw_flatten <- apply(this_draw,2,sum)/sum(this_draw)
  sim_votes_max <- rbind(sim_votes_max, this_draw_flatten)
}

sim_votes_diff <- sim_votes_max - sim_votes_true


plot("n", 
     main = "Estimated differnce in share of votes for each party\n(No distance condition - true distance condition)", 
     ylim = c(-0.5,0.5),
     xlim = c(1,14),
     xlab ="Party",
     ylab = "Percentage point difference",
     xaxt="n")
axis(side=1, at=1:14, labels = c("NA","A","B", "C", "F", "H", "I", "M", "O", "V", "Æ", "Ø", "Å", "U"))
for (i in 1:30){
  lines(100*sim_votes_diff[i,])
}
lines(c(1,14),c(0,0),lty=2, col="red")
reduced_voters %>% colnames()
credible_intervals <- 100*apply(sim_votes_diff,2,HPDI,0.95)
max(credible_intervals[2,]) #very certain that no party would have gotten 0.08 percentage points more of vote share
min(credible_intervals[1,]) #or lost more than 0.1 perentage points.
```
No clear bias, and all simulations agree that any difference is certainly smaller than 1 percentage point.

### blue vs red block
```{r}
blue_parties <- c(4,6,7,9,10,11)
red_parties <-  c(2,3,5,12,13)

plot("n", xlim=c(0.435,0.51), ylim = c(1,N_samples), 
     main = "Total vote share for Red and Blue block for each posterior draw\n0 DtPS", 
     ylab = "Posterior Draws", 
     xlab = "Percentage of votes",
     yaxt="n")
lines(apply(sim_votes_max[1:N_samples,red_parties],1,sum)/apply(sim_votes_max[1:N_samples,c(-1)],1,sum),1:N_samples, col = "red")

lines(apply(sim_votes_max[1:N_samples,blue_parties],1,sum)/apply(sim_votes_max[1:N_samples,c(-1)],1,sum),1:N_samples, col = "blue")
lines(c(0.5,0.5),c(0, N_samples+1),lty=2)
```


